\chapter{Notes}\label{sec:notes}

\section{Grammars}\label{grammars}

Equivalent: surrounded by context, length left \(\leq\) length right,
\href{https://en.wikipedia.org/wiki/Kuroda_normal_form}{Kuroda normal
Form}

\section{Summary}\label{summary}

\(FO[t(n)]\):
\(A \in S \Leftrightarrow A \models \left([QB]^{t(|A|)}M_{0}\right)(\overline{c}/\overline{x})\)

\noindent\(VAR[v + 1] = \bigcup_{c = 1}^{\infty}FO-VAR[2^{cn^{v}}, v + 1]\):
Unbounded iterations, using at most \(v + 1\) vars

\noindent\(ITER[\text{arity } k]\): Simultaneous iterations of relations of arity
k as in LFP (no monoticity requirement)

\noindent\(VAR[k + 1, r] / ITER[\text{arity } k, r]\): Same plus restricted
variables of total bits \(r + \mathcal{O}(1)\)

\noindent\(DSPACE[s(n)] = VAR\left[ k + 1, \log\left( \frac{s(n)}{n^{k}} \right) \right] = ITER[\text{arity } k, \log\left( \frac{s(n)}{n^{k}} \right)]\)
with \(k = \left\lfloor \log_{n}(s(n)) \right\rfloor\) Problem for
generalisation: For \(C_{t}(\overline{x}, \overline{b})\) (tape cell
\(\overline{x}\) at time \(t\) is \(\overline{b}\)), for nondeterminism
we could have the same, but other tape cells are not independent.

\noindent\(FO-VAR[s(n), v(n)]\): \(VAR[s(n)]\) + constant number of
\(v(n)\log(n)\) bit variables, which can only be tested by \(BIT\).

\noindent\(SO[t(n), \text{arity }k]\): \(FO[t(n)]\), but with quantification over
relation variables of arity at most k

\noindent\(NSPACE[n^{k}] = SO(\text{arity }k)(TC) \subseteq SO(\text{arity } k)[n^{k}]\)

All formulas of \(SO(\text{arity }k)(TC)\) can be written in the form \[
(TC_{struct_{1}, struct_{2}}\alpha)(\overline{false}, \overline{true})
\] where \(\alpha\) is quantifier-free and both structs have at most
\(O(n^{k})\) bits


\[SO(\text{arity } k)[t(n)] = FO-VAR\left( t(n), \frac{n^{k}}{\log(n)} \right)\]

\[\text{DSPACE}[s(n)] \subseteq \text{NSPACE}[s(n)] \subseteq \text{FO-VAR}\left[ s(n), \frac{s(n)}{\log(n)} \right] \subseteq ATIME[s(n)^{2}]\subseteq \text{DSPACE}[s(n)^{2}]\]
Proof works by looking at alternating time paths of length \(2^{r}\) in
the \(NSPACE\) computation graph.



\section{\texorpdfstring{Extending
\(C_{t}(\overline{x}, \overline{b})\)}{Extending C\_\{t\}(\textbackslash overline\{x\}, \textbackslash overline\{b\})}}\label{extending-c_toverlinex-overlineb}

Adding an additional variable for non-deterministic choices
\(C_{t}(\overline{c}, \overline{x}, \overline{b})\), where
\(\overline{c}\) are the non-deterministic choices, makes the states
size grow exponentially in \(O(2^{cn^{k}})\) (one for each move of the
NTM). So we have way more memory then \(ITER[k + 2]\), which makes it
strictly more powerful. We need to have all the choices as we can only
inductively define \(C_{t}\) using the previous related choices.

\section{\texorpdfstring{Direct transformation of
\(\text{SO}(\text{arity }k)(TC)\) to
\(\text{FO-VAR}\left[ 1, \frac{n^{k}}{\log n} \right](TC)\)}{Direct transformation of SO(\textbackslash text\{arity \}k)(TC) to FO-VAR\textbackslash left{[} 1, \textbackslash frac\{n\^{}\{k\}\}\{\textbackslash log n\} \textbackslash right{]}(TC)}}\label{direct-transformation-of-sotextarity-ktc-to-fo-varleft-1-fracnklog-n-righttc}

\(SO(\text{arity }k)(TC) \subseteq FO-VAR\left[ 1, \frac{n^{k}}{\log n} \right](TC)\):
We know that any formula in \(SO(\text{arity }k)(TC)\) can be written in
the form \[
(TC_{struct_{1}, struct_{2}}\alpha)(\overline{false}, \overline{true})
\] where \(\alpha\) is quantifier-free. Using the equivalence
\(BIT(Y, \overline{x}) \Leftrightarrow Y(\overline{x})\), we can write
anything in \(\alpha\) using
\(FO-VAR\left[ 1, \frac{n^{k}}{\log n} \right]\). Any relation in
\(struct_{1}, struct_{2}\) can also be rewritten using the extended
variables. As we now any Relation with \(false\) is converted to \(0\)
and any relation with \(true\) is converted to \(1\), we have
\((\overline{false}, \overline{true})\) that can now be represented by
\((\overline{0}, \overline{max})\).

\(SO(\text{arity }k)(TC) \supseteq FO-VAR\left[ 1, \frac{n^{k}}{\log n} \right](TC)\):
We know that any formula in \(FO(TC)\) can be written as
\((TC_{struct_{1}, struct_{2}}\alpha)(\overline{0}, \overline{max})\).
We can extend this result to include the extended variables. For the
usage of the extended variables in \(BIT\), we have an atomic formula,
so this is trivial. For quantification, we can handle it the same way as
normal quantification with the modification of the ``counting'' variable
with a tuple of \(k\) counting variables, and the successor relation on
these \(k\) variables. We haven't shown yet that negation can be handled
by this. In the proof that \(FO(pos~TC) = FO(TC)\), we only ask for a
successor relation on the vertices (consisting of the extended variables
and normal variables), which we assume from the beginning, so this also
holds for
\(FO-VAR\left[ 1, \frac{n^{k}}{\log n} \right](pos~TC) = FO-VAR\left[ 1, \frac{n^{k}}{\log n} \right](TC)\).
Now, we can do the same as above for the other direction to convert a
formula of the form
\((TC_{struct_{1}, struct_{2}}\alpha)(\overline{0}, \overline{max})\) to
\((TC_{struct_{1}, struct_{2}}\alpha)(\overline{false}, \overline{true})\).

\section{\texorpdfstring{Can we prove that there is no way to use
\(TC\) with a \(FO\) language that has less then \(O(n^{k})\)
variables?}{Can we prove that there is no way to use TC with a FO language that has less then O(n\^{}\{k\}) variables?}}\label{can-we-prove-that-there-is-no-way-to-use-tc-with-a-fo-language-that-has-less-then-onk-variables}

We have Arity theorems
\href{https://pdf.sciencedirectassets.com/271596/1-s2.0-S0168007200X00292/1-s2.0-0168007295000720/main.pdf?X-Amz-Security-Token=IQoJb3JpZ2luX2VjEA0aCXVzLWVhc3QtMSJHMEUCIQCZEfzPpPqb83EtF\%2BLgRMsETvrj14M6SqdN3VHcNPEkdAIgWF\%2BnohPDt8o\%2FXZc7b7P3HqFm72ZMaOoAg4MLrtTMUtYquwUIpv\%2F\%2F\%2F\%2F\%2F\%2F\%2F\%2F\%2F\%2FARAFGgwwNTkwMDM1NDY4NjUiDOiLVE0art2rY62EliqPBQbQSd\%2FAcBktlD2Axqi2hf20al4TnosyDnMOkEbqwIQAyET8IMr\%2B776j5WCIdUO85KSby8IvsXcoMaeJllnCTWiAtcQOCv0tjt1Rome20\%2B50\%2BMDR2DM6BHnHFA3gOFm3PAoX427r\%2BysNLJl2dbF4YAXE2Rb02hYUm47FnV\%2FiV6LDZgVk56JvT2yASPDa1TYaBTCUt09gepbCkOkC1Tzac2OEzReNX5RGnSGRJRwZN7QDFmHMEhiE1YO1sYOGJgnrWvjBNlgRGe66ZDXA6H9FlnP3KfXpA\%2BRlBp9SfqipL6iEj8iTLlY8Iw4Lx8SLSR4WwHt5ZSopsZj1SZlTKtBkaHqb4\%2ByS8iYsy7JUiFML1LvIE4ttxy7fM6qY5XNU0NSNb6NF\%2FStuygMbc5kU3Xs5D\%2BYtei30HQKStUmol3nJYpBbgTpaUCPZTWF1xfFeF2r0rHtT8SGvWCZNI0N78o6YYf\%2FPiFrC\%2FLb\%2BaGc0DtkdqrLP3L3Rrf2cIXsxqZ4LwlEdj6fEIKZxw71c95o1h1PlnB72qAg62JKOdNtHM3Gpw7yeuU1kAdNEctOU2WMvhqqjDHlcerDbspa39NRMVIgrKI0wHhphT42iSbZC2rVP7cgpqBFADJqrT1C3ROwrQngxGDjCqpaDgR3Lh\%2BFCvJ6Hj6aficQkrXnZKYPd2ZxF0LoEw5F0K1eEAoi\%2F31m3Wd8MhY\%2FzlDt5Owh4zIlIsv9AEG0byN\%2F\%2FEs30PT469\%2BiRlOlCAJ0iYqpvsVk3whvBPEHae8O3LFUG7gWFK50eQ\%2FFQ9KVcy0sLMCNpK286rnXOYvU18pdah3GbEt\%2BG642xb\%2F3y\%2BPNG6oMZ\%2B485zsKgusLfgILd7\%2FU7w4E8yeE8VrHFPFcw3\%2B2wswY6sQFACa99Hzo8imVILu9U2u8E7F2wa2uOniA6UidGdcybvYvso0UTqK53Ki\%2FnJ5tT6ADTd52VSNtgviQbNSxdnAdsifHTNGD4xTutf9zwfV1AVAWosXcbKwC\%2BGd1Cmoh4JMrd4uMlW2FtCpKGbmKW9pcJkmjNRHacw3IN\%2FC\%2F\%2BI4yRBQZyBwoKevOICAIRcr8XsVMxvpSFg4M6Amz3UbgG1Tf5\%2FaSW\%2BAn2\%2BHYGkfJuV5Io8OU\%3D&X-Amz-Algorithm=AWS4-HMAC-SHA256&X-Amz-Date=20240614T123128Z&X-Amz-SignedHeaders=host&X-Amz-Expires=300&X-Amz-Credential=ASIAQ3PHCVTYVAGA3OF6\%2F20240614\%2Fus-east-1\%2Fs3\%2Faws4_request&X-Amz-Signature=42e03b07238e340e4071f6b533897676385912b01735ff90b4ce8ee3841179f2&hash=a3824b39f0ff05985504a3e08bf09f2572a737ad6204bcb9d1f18f599e123db5&host=68042c943591013ac2b2430a89b270f6af2c76d8dfd086a07176afe7c76c2c61&pii=0168007295000720&tid=spdf-86d73916-b662-4ed3-b6de-753dfc8128db&sid=9e433d8d71f1564cc768be65fe1f9e328634gxrqb&type=client&tsoh=d3d3LnNjaWVuY2VkaXJlY3QuY29t&ua=070f5e5851065a5958&rr=893a566aeef5b39e&cc=ch}{Arity
hierarchies Martin Grohe}.
Is this applicable because ordered
structure? Diagonalisation in logic?

FO{[}\(2^{n}\){]}, SO{[}n{]} can both express REACH\(_{dl}\)

\emph{Claim:} If extended variables are allowed everything, nothing
changes.

\emph{Define} Superrelations, using the extended variables. Does not
work because input to big Not needed as we are working on string
structures

If we allow extended variables for relations we have undefined for some,
and if we check, we could ``copy'' bits into FO variable Numerical
relations \(\leq, SUC(X, Y), +, \times\) are definable via BIT anyways.

If no further relations are added, we have already that they reflect
exactly some NSPACE complexity class, so we can not do better.

\textbf{Approach 2} We have that the FO{[}1, s(n)/log(n){]}(TC)
correspond exactly to our complexity classes. The Space hierarchy tells
us that we can not do better while using these constructions.

Now assume we add iterated quantifier blocks. We have multiple ways of
doing this. We can have the TC inside or outside

\[
TC_{s(n)}([QB]^{t(n)}\varphi) \equiv TC_{O(n)}(\phi )
\] In both cases, we can replace the TC operator with a quantifier block
iterated \(o(n)\) times. So we have formulas of the form

\[
[QB_{1}]^{s(n)}[QB_{2}]^{t(n)} \psi
\] We can include a counting variable of which counts to \(o(n)\) and
then switches block to get only one combined block.

so we then have a formula in FO{[}s(n) + t(n), s(n)/log(n){]}, which is
also in FO{[}t(n), s(n)/log(n){]}(TC) or FO{[}s(n), s(n)/log(n){]}(TC)
(or TC{[}FO{[}s(n), s(n)/log(n){]}{]}, but does just not use the TC
operator. So if s(n) \(\not\in o(n)\) then this is equivalent or more
difficult then showing that
\(SO(\text{arity } k)[n^{k}] = SO(\text{arity } k)(TC)\) Otherwise, if
\(t(n) \geq s(n)\) we have to show that
\(FO\left[ t(n), \frac{s(n)}{log(n)} \right] = SO(\text{arity } k)(TC)\).
The only problem arises if \(s(n) > t(n)\), as then the generated new
formula is not in the original class. What would happen is that then
\(SO(arity~k)(TC) = FO\left[ t(n), \frac{s(n)}{\log(n)} \right](TC)\subseteq FO\left[ s(n), \frac{s(n)}{\log n} \right] \subseteq SO(\text{arity } k)[n^{k}]\)
We know that \(s(n)\) must be in \(\Omega(\sqrt{ n })\) by Savitch's
theorem. Also if \(s(n) \in o(n)\), we would have a stronger result then
Savitch's theorem as then \(NSPACE[n] \subseteq DSPACE[s(n)^2]\), which
is an open problem since almost 50 years.

In the end, this proves that combining TC and quantifier iterations is
as difficult either as showing an improvement of Savitch's theorem or is
equivalent to having an alternative characterisation using only iterated
quantifiers.


\section{Generalised quantifiers}\label{generalised-quantifiers}

By \href{https://link.springer.com/article/10.1007/bf01268616}{double
arity} we have FO(TC) = FO(\(Q_{TC}\)) which is the tc quantifier. The
tc Quantifier has type \(\langle 2, 1 ,1 \rangle\) and
\(k\)-vectorisation to make it generalised to vectorisations. There is a
theorem which states that no family of quantifiers of arity \(<2k\)
suffices to express these properties. So We can effectively say that
they are the easyest.

This is true for FO-generalised quantifiers on unordered structures, but
unclear for \href{https://core.ac.uk/download/pdf/81931656.pdf}{SO-gq}

Problem is that we can query bits, so we have even more than just an
order

Can we port the results directly?

\section{\texorpdfstring{FO\(\exists\){[}n{]}}{FO\textbackslash exists{[}n{]}}}\label{foexistsn}

\begin{description}

\item[FO\(\exists\){[}s(n){]}]
A set \(S\subseteq \text{STRUCT}[\tau]\) is in FO\(\exists\){[}s(n){]}
if and only if there exists formulas \(M_{i}, N_{j}\),
\(0 \leq i \leq k, 1 \leq j \leq r\) from \(\mathcal{L}(\tau)\), a tuple
\(\overline{c}\) of constants and a quantifier block \[
QB = [(\exists x_{1}.M_{1})(\forall b_{1}.N_{1})\dots(\exists x_{k}.M_{k})(\forall b_{r}.N_{r})]
\] such that for all \(\mathcal{A} \in \text{STRUCT}[\tau]\), we have \[
\mathcal{A} \in S \Leftrightarrow \mathcal{A} \models ([QB]^{s(\left| \mathcal{A} \right|)}M_{0})(\overline{c} / \overline{x})
\]
\end{description}

FO\(\exists\)-VAR{[}t(n), s(n){]} is the same with extended variables

We can write TC as

\[
\begin{aligned}
QB \equiv~& (\forall b_{1}.M_{1})(\exists\overline{Z}) (\forall b_{2})(\exists \overline{A}, \overline{B}.M_{2})(\exists\overline{X}, \overline{Y}.M_{3}) \\
M_{1} \equiv~& \neg(\forall \overline{z} (\text{BIT}(\overline{X}, \overline{z}) \leftrightarrow \text{BIT}(\overline{Y}, \overline{z})) \lor \varphi(\overline{X}, \overline{Y})) \\
M_{2} \equiv~&(b_{2} \land \forall \overline{z} (\text{BIT}(\overline{X}, \overline{z}) \leftrightarrow \text{BIT}(\overline{A}, \overline{z}))\land \forall \overline{z} (\text{BIT}(\overline{B}, \overline{z}) \leftrightarrow \text{BIT}(\overline{Z}, \overline{z}))) \lor \\
&(\neg b_{2} \land \forall \overline{z} (\text{BIT}(\overline{Z}, \overline{z}) \leftrightarrow \text{BIT}(\overline{A}, \overline{z}))\land \forall \overline{z} (\text{BIT}(\overline{B}, \overline{z}) \leftrightarrow \text{BIT}(\overline{Y}, \overline{z}))) \\
M_{3} \equiv~&\forall \overline{z} (\text{BIT}(\overline{X}, \overline{z}) \leftrightarrow \text{BIT}(\overline{A}, \overline{z}))\land \forall \overline{z} (\text{BIT}(\overline{B}, \overline{z}) \leftrightarrow \text{BIT}(\overline{Y}, \overline{z}))
\end{aligned}
\]

And then
\([QB]^{cs(n)}(false)(\overline{false} / \overline{X}, \overline{true} / \overline{Y}) \equiv (TC_{\overline{X}, \overline{Y}}\varphi(\overline{X}, \overline{Y}))(\overline{false}, \overline{true})\)

Thus, we have NSPACE{[}s(n){]} \(=\) FO-VAR{[}1, s(n)/log(n){]}(TC)
\(\subseteq\) FO\(\exists\)-VAR{[}s(n), s(n)/log(n){]}

For the reverse containment, we can simulate the formula with a
NSPACE{[}s(n){]} machine by existentially guessing the existentially
quantified variables and remembering the universal choices in \(s(n)\)
space. Does not work because we need some backtracking stuff, and thus
remembering the existential vars.

Also,
\(FO-VAR[1, s(n)/log(n)](TC) \subseteq FO-VAR\left[ \frac{s(n)}{\log(r(n))}, \frac{s(n)r(n)}{\log(n)} \right]\)
by doing multiple steps at once, and universal quantification over
\(r(n)\) bit variables only.

\section{Alternating machines and FO-VAR}\label{sec:somewhat-unrelated-stuff}

\(ATIME-SPACE[s(n)^2, s(n)] \subseteq FO-VAR\left[ s(n)^2, \frac{s(n)}{\log(n)} \right]\)

We have a state space of \(2^{\mathcal{O}(s(n))}\), so we need only
\(s(n)\) bit Variables to encode it. We can assume that there exist
formulas \(existencial(\overline{X})\) which is true if \(\overline{X}\)
is an existential state, and one \(\varphi(\overline{X}, \overline{Y})\)
telling us that we can transition from \(\overline{X}\) to
\(\overline{Y}\).

Then, we have \[
\begin{aligned}
QB \equiv~&(\forall b.M_{1})(\exists\overline{A}.\varphi(\overline{A}, \overline{X}))(\forall \overline{B}.M_{2})(\exists \overline{X}. \forall z(\text{BIT}(\overline{X}, \overline{z}) \leftrightarrow \text{BIT}(\overline{B}, \overline{z}))) \\
M_{1} \equiv~&\neg\forall z(\text{BIT}(\overline{X}, \overline{z}) \leftrightarrow \text{BIT}(\overline{Y}, \overline{z})) \\
M_{2} \equiv~& (existential(\overline{X}) \to \forall z(\text{BIT}(\overline{A}, \overline{z}) \leftrightarrow \text{BIT}(\overline{B}, \overline{z}))) \land \varphi(\overline{X}, \overline{B})
\end{aligned}
\] and \[
\left([QB]^{s(n)^2}(false)\right)(\overline{false} / \overline{X}, \overline{true} / \overline{Y})
\]

\[
    \begin{aligned}
        ATSR[t(n), s(n), r(n)] &\subseteq FO-VAR\left[ r(n)\log\left( \frac{t(n)}{r(n)} \right), \frac{s(n)}{\log n} \right] \\ &\subseteq ATSR\left[r(n)s(n)\log\left(\frac{t(n)}{r(n)}\right), s(n), r(n)\log\left( \frac{t(n)}{r(n)} \right)\right]
    \end{aligned}
\]
The second containment is a consequence of how we simulate the formula
with the alternating Turing machine, as we have
\(FO-VAR\left[ a(n), \frac{s(n)}{\log n} \right] \subseteq ATSR[a(n)s(n), s(n), a(n)]\)
by using the normal construction in Savitch's theorem.

In the first containment, we try to use the method of simulating NSPACE
for each alternation. First, we show that \[
\sum_{j = 0}^{r(n)}\log(a_{j}) \leq r(n)\log\left( \frac{t(n)}{r(n)} \right)
\] under the condition that \(\sum_{j= 0}^{r(n)}a_{j} = t(n)\).

This can be shown using Jensen's inequality. \[
\frac{\sum_{j = 0}^{r(n)}\log(a_{j})}{r(n)} \leq \log\left( \frac{\sum_{j= 0}^{r(n)}a_{j}}{r(n)} \right) = \log\left( \frac{t(n)}{r(n)} \right)
\] as \(\log\) is concave. Multiplying with \(r(n)\) gives the claim.

\(Y\) accepting \(X\) start \(I\) actual intermediate end \(S\) actual
start \(E\) actual end \(A, B\) copy vars \(b_{path}\) if we are trying
to find or not find a path \[
\begin{aligned}
QB \equiv&~(\forall b_{0}.M_{1})(\exists b_{0}.N_{1})(\exists b_{1}.M_{2})(\exists Z_{e}.M_{3})(\forall Z.N_{2})\\
&~(\exists b_{2e})(\forall b_{2}.N_{3})(\exists A, B.M_{4})(\forall S, E.M_{5})(\exists A, B.M_{6})(\exists I, X.M_{7})(\exists b_{0}.N_{4})(\exists b_{path}.N_{5}) \\
M_{1} \equiv&~b_{path} \to \neg(I = Y \land (S = E \lor \varphi(S, E))) \\
N_{1} \equiv&~\neg b_{path} \to \neg(S = E \lor \varphi(S, E)) \\
M_{2} \equiv&~\neg b_{1} \leftrightarrow (S = E \lor \varphi(S, E)) \\
M_{3} \equiv&~existential(Z_{e}) \leftrightarrow existential(X)\\
N_{2} \equiv&~b_{path} \to Z = Z_{e}\\
N_{3} \equiv&~ \neg b_{path} \to b_{2} = b_{2e}\\
M_{4} \equiv&~(b_{1} \land ((\neg b_{2} \land A = S  \land B = Z) \lor (b_{2} \land A = Z \land B = E))) \lor \\
&~(\neg b_{1} \land  A = I \land (B = Y \lor (existential(B) \leftrightarrow \neg existential(I)))) \\
M_{5} \equiv&~S = A \land ((( b_{1} \lor existential(S)) \land E = B) \lor \\
&~(\neg( b_{1} \lor existential(S)) \land (E = Y \lor (existential(E)\leftrightarrow \neg existential(S))))) \\
M_{6} \equiv&~(\neg b_{1}  \land S = A \land E = B) \lor (b_{1} \land X = A \land I = B)\\
M_{7} \equiv&~X = A \land I = B \\
N_{4} \equiv&~b_{0} = b_{path} \\
N_{5} \equiv&~b_{1} \to b_{path} = b_{0}
\end{aligned}
\]

In this formula, we have
\(\left([QB]^{r(n)\log(t(n)/r(n))}(\neg b_{path})\right)(true / b_{path}, AS / S, X, I, S, E, AE / Y)\)
is true if and only if the machine accepts.

\(\implies FO-VAR\left[ a(n), \frac{s(n)}{\log n} \right] \subseteq ATSR[a(n)s(n), s(n), a(n)] \subseteq FO-VAR\left[ a(n)\log(s(n)), \frac{s(n)}{\log n} \right]\)
